HO unification from object language to meta language

### 2024-02-27 paper/main.tex
4

% \input{lang_descr.tex}

% Description of languages (OL and ML)

\def\eqfo{eq\_fo\xspace}

In order to reason about unification of the terms of the objet language within the
meta language, we start by formally describing the two languages. Employing
meta-programming for this purpose, we present in \cref{code:common-terms} a new type for the terms of
the OL.
%
\setlength{\abovecaptionskip}{0pt}
\setlength{\belowcaptionskip}{-13pt}

\begin{figure}[H]
  \begin{tabular}{ll}
  \begin{minipage}{0.21\textwidth}
  \begin{elpicodetab}
kind fm type.
type fapp list fm -> fm.
type flam (fm -> fm) -> fm.
type fcon string -> fm.
type fuva nat -> fm.
\end{elpicodetab}
  \end{minipage}
  &
  \begin{minipage}{0.24\textwidth}
  \begin{elpicodetab}
kind tm type.
type app list tm -> tm.
type lam (tm -> tm) -> tm.
type con string -> tm.
type uva nat -> list tm -> tm.
  \end{elpicodetab}
  \end{minipage}
  \end{tabular}
  \caption{Source and Target language}
  \label{code:common-terms}
  \Description[code:common-terms]{code:common-terms}
\end{figure}

\noindent
This encoding is very similar to the one introduced in \cref{sec:intro},\todo{say we do not care about types} except
for the \elpiIn{all} constructor. We explicitly neglect it since blabla.
Moreover, since we are working with unification variables we need to introduce a
new constructor.

In the case of the OL unification variables \elpiIn{fuv} have no scope.
For example the statement of the instance \coqIn{forall_dec} contains the
term \coqIn{P x} where \coqIn{P} stands for a function of type
\coqIn{A ~$\to$~ Prop}. However our representation of \coqIn{P x} in the meta
language is \elpiIn{app[P, x]} where \elpiIn{P} is of type \elpiIn{ftm}.
In the terms of the ML unification variables come with a scope (the second
argument of the \elpiIn{uva} constructor).
Of course some attention
should be payed when dealing with \elpiIn{uva}, since we have to certify
that an \elpiIn{uva i} remain in the pattern fragment, that is, the
list of term in the scope of \elpiIn{i} is always a list of distinct names.

Unification variables of the OL are identified by a natural number
referring to memory address. 
Finally, the memory of
the OL is depicted by the following type abbreviation:

\marginpar{maybe move code to the appendix}
\begin{elpicode}
typeabbrev memory A (list (option A)).
type set nat -> memory A -> A -> o.
set N S T :- nth N S (some T).
type unset nat -> memory A -> o.
unset N S :- nth N S none.
type assign nat -> memory A -> A -> memory A -> o.
assign z [none|M] T [some T|M].
assign (s N) [X|M] T [X|M1] :- assign N M T M1. 
\end{elpicode}

\noindent 
which is a list of optional terms. If a cell is \elpiIn{none}, then the variable
corresponding to this cell is not instantiated.

Another important implementation should be provided in order to represent the
lambda abstractions and the applications of the ML. As outlined in
\cref{sec:encodings}, <<\elpiIn{x\ f x}>> has not equal to 
<<\elpiIn{lam x\ app[f, x]}>>, since, even though the two terms represent the same
concept, the latter uses the abstraction and application of the OL.

@@

\noindent
In particular, the node \elpiIn{abs} stands for the lambda abstraction of the ML.
The
node \elpiIn{val} contains terms of the object language. The node for variable
application is not really necessary, since, using the syntax of the ML, it is
defined by putting two terms of the ML side by side. For example, the terms 
<<\elpiIn{x\ f x}>> becomes <<\elpiIn{abs x\ f x}>>.

The memory of the ML is represented by a \elpiIn{subst}.

\noindent
with the invariant that the \elpiIn{tm} inside the cell never contains the node
\elpiIn{fo_uv}. Note that, reciprocally, the \elpiIn{fo_subst} does not contain
any node of the form \elpiIn{ho_uv}.

>>>>

In order to reason about unification we provide a description of the
\Fo{} and \Ho languages where unification variables
are first class terms, i.e. they have a concrete syntax. We keep these languages
minimal, for example we omit the \elpiIn{all} quantifier of DTT we used
in the example in Section~\ref{sec:intro}. 
%
\setlength{\abovecaptionskip}{0pt}
\setlength{\belowcaptionskip}{-13pt}

\begin{figure}[H]
  \begin{tabular}{ll}
  \begin{minipage}{0.21\textwidth}
  \begin{elpicodetab}
kind fm type.
type fapp list fm -> fm.
type flam (fm -> fm) -> fm.
type fcon string -> fm.
type fuva nat -> fm.
\end{elpicodetab}
  \end{minipage}
  &
  \begin{minipage}{0.24\textwidth}
  \begin{elpicodetab}
kind tm type.
type app list tm -> tm.
type lam (tm -> tm) -> tm.
type con string -> tm.
type uva nat -> list tm -> tm.
  \end{elpicodetab}
  \end{minipage}
  \end{tabular}
  \caption{\Fo{} and \Ho language}
  \label{code:common-terms}
  \Description[code:common-terms]{code:common-terms}
\end{figure}

\noindent
In the case of \Fo{} unification variables \elpiIn{fuva} have no explicit scope:
Unification variables standing for functions are applied to variables
via the \elpiIn{fapp} constructor.
For example in the statement of the instance \coqIn{forall_dec} the
term \coqIn{P x} is represented as \elpiIn{fapp[fuva N, x]}, where \elpiIn{N}
is of type \elpiIn{nat} and \elpiIn{x} is a bound variable.

In \Ho the representation of \coqIn{P x} is instead \elpiIn{uva N [x]}.
We say that the unification variable \elpiIn{uva N L} is in \llambda iff
\elpiIn{distinct-names L} holds.

\begin{elpicode}
type distinct-names list tm -> o.
distinct-names [].
distinct-names [X|XS] :- name X, not(mem X XS),
  distinct-names XS.
\end{elpicode}

\noindent
The \elpiIn{name} builtin predicate tests if a term is a bound variable.

In both languages unification variables are identified by a natural number,
which can be seen as a memory address. The memory and its associated
operations are described below:

\marginpar{maybe move code to the appendix and keep only signatures}
\begin{elpicode}
typeabbrev memory A (list (option A)).
type set nat -> memory A -> A -> o.
set N S T :- nth N S (some T).
type unset nat -> memory A -> o.
unset N S :- nth N S none.
type assign nat -> memory A -> A -> memory A -> o.
assign z [none|M] T [some T|M].
assign (s N) [X|M] T [X|M1] :- assign N M T M1. 
\end{elpicode}

\noindent 
If a memory cell is \elpiIn{none}, then the corresponding unification variable
is not set.

Since \Ho unification variables have a scope, their solution needs to be
abstracted over these in order to enables the instantiation of a single
solution to different scopes. On the contrary solutions to \Fo variables
are just terms.

@@

\noindent
One can see \elpiIn{abs} as a syntax for the lambda abstraction of the meta
language.
%Luckily  we need to reason on these abstractions only when we
%manipulate the solution given by the unification of the meta language; when
%we manipulate the terms we don't 
%
%Another important implementation should be provided in order to represent the
%lambda abstractions and the applications of the ML. As outlined in
%\cref{sec:encodings}, <<\elpiIn{x\ f x}>> has not equal to 
%<<\elpiIn{lam x\ app[f, x]}>>, since, even though the two terms represent the same
%concept, the latter uses the abstraction and application of the OL.
%
%In particular, the node \elpiIn{abs} stands for the lambda abstraction of the ML.
% The
% node \elpiIn{val} contains terms of the object language. The node for variable
% application is not really necessary, since, using the syntax of the ML, it is
% defined by putting two terms of the ML side by side. For example, the terms 
% <<\elpiIn{x\ f x}>> becomes <<\elpiIn{abs x\ f x}>>.

The memory \Fo{} is called \elpiIn{fsubst} while the one of \Ho is
called \elpiIn{subst}.
Both have the invariant that they are not cyclic, TODO explain.
